---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
In order to complete the assignment, we are going to need some libraries to be 
loaded into the R environment.  

1. lubridate
2. dplyr
3. ggplot2

We shall also set some default variables so we can use them throughout the 
exercise.

```{r loadpackages, message=FALSE, warning=FALSE}
# Required packages
packages <- c('lubridate', 'dplyr', 'ggplot2')

# Default variables
dataURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
dataFolder <- file.path(getwd(), 'data/')
filename <- 'activity.zip'
extractedFile <- 'activity.csv'

# Checking packages are downloaded and are loaded in environment.
missingPackages <- packages[(!(packages %in% installed.packages()))]
if(length(missingPackages) > 0)
  install.packages(missingPackages)

if(!all(unlist(lapply(packages, require, character.only = TRUE))))
  stop('Cannot install packages. Quitting...')

```
We also need some functions to help with the processing, namely:

* checkPackages() - to load required packages into the R environment.
* loadData() - to download and load the data into the R environment.
* cleanData() - to clean the data as part of the data analysis.

Let's set these up now...
```{r helperfunctions}

# Function to download and load dataset. Returns loaded dataset
loadData <- function() {
  fullFileName <- file.path(dataFolder, filename)
  
  dir.create(dataFolder, showWarnings = FALSE)
  #  download.file(dataURL, fullFileName)
  unzip(zipfile = fullFileName, exdir = dataFolder, files = extractedFile)
  
  # Return dataset
  tbl_df(read.csv(file.path(dataFolder, extractedFile)))
}

# Function to clean the loaded data set and to specification. Returns 
# cleaned data set 
cleanData <- function(data_file, valueOfNASteps) {
  # Clean up the dates usling lubridate and replace NA in steps field
  # with value in valueOfNASteps argument
  
  data_file$date <- ymd(data_file$date)
  if(!missing(valueOfNASteps))
    data_file$steps <- replace(
      data_file$steps, which(is.na(data_file$steps)), 
      valueOfNASteps)
  
  data_file
}

```
With the above done, we need to load the data set into the environment and make
it ready to answer the exercise questions. Let's load the original data set, 
and a cleaned data sets that sets the **NA** values of **step** to the value
**zero**.

```{r loadDataSets}

# Loading the original data set into the dataSet variable
dataSet <- loadData()

# Cleaning steps in original data set by replacing NA with 0
dataFile <- cleanData(dataSet, 0)

```
So with the preprocessing done, let's try and answer some of the questions.

## 1. What is mean total number of steps taken per day?
Having set the **NA** values to **zero**, we have the total number of steps 
as `r as.integer(sum(dataFile$steps))` steps. 

We can calculate the mean number of steps for each day as below: 

```{r meanDailySteps }
totalsteps <- dataFile %>% group_by(date) %>% 
  summarise(total = sum(steps, na.rm = TRUE))

data.frame(totalsteps) # To print out full list.
```
We can also make a __histogram__ of the above totals to make it easier to 
visualise: 

```{r averageStepsHistogram }
totalsteps %>% ggplot(aes(total)) + geom_histogram(binwidth = 350) + 
  ggtitle('Histogram of total number of steps')
```
Let's also calculate the mean and median for each day and save it. It may come 
in handy later on...

```{r summaryData}
summaryData <- dataFile %>% group_by(date) %>% 
  summarise(total = sum(steps), 
            mean = mean(steps), 
            median = median(steps))

data.frame(summaryData)
```
## 2. What is the average daily activity pattern?

Let's have a look at the average daily activity pattern over time...

```{r activityPattern}
ggplot(summaryData, aes(date, mean)) + 
  geom_line() + 
  labs(title = 'Time Series Plot', x = 'Date', y = 'Average Step Count')
```

From the graph, we can see that the maximum numebr of step counts occurs on 
`r summaryData[which.max(summaryData$mean), ]$date`.

## 3. Imputing missing values
The data shows that there are `r sum(is.na(dataSet))` number of **NA** 
step values.

To fill in these values, we shall use the daily means. We shall merge the 
data set with the summary means and then replace the **NA** values with the 
respective means to create a new cleaned data set:

```{r mergingCleaning }
mrg <- merge(cleanData(dataSet), summaryData, by = 'date')
NADataSet <- cleanData(dataSet)
indices <- which(is.na(NADataSet$steps))
NADataSet$steps <- replace(NADataSet$steps, indices, mrg$mean[indices])

```

So how does the new data set look like?
```{r }
NADataSet %>% group_by(date) %>% 
  summarise(
    totalsteps = sum(steps),
    mean = mean(steps), 
    median = median(steps)) %>%
  data.frame
```

Demonstrating the number of steps taken each day on a histogram gives:

```{r } 
ggplot(NADataSet, aes(steps)) + geom_histogram()
```

## 4. Are there differences in activity patterns between weekdays and weekends?
To answer this, will first have to determine which of the days in the data
set are on weekdays and weekends, and then plot the data on a line graphs - 
weekend and weekday.

So, let's see how to do this.
```{r weekGraph }
# Let's identify which days make the weekend.
weekend <- c('Sat', 'Sun')

# And set the classifiers for marking each day
title <- c('Weekend', 'Weekday')

NADataSet %>%
  
  # Merge the data set with the Summary Data set.
  merge(summaryData, by = 'date') %>% 
 
   # And mark each day as weekened or weekday. We add 1 as FALSE translates
  # to 0 and 1 translates to TRUE, but vector indices begin from 1.
  mutate(weekday = title[weekdays(date, TRUE) %in% weekend + 1]) %>% 
  
  # Summarise the data by grouping and averaging the steps by interval and day
  group_by(weekday, interval) %>% 
  summarise(AverageSteps = mean(steps)) %>% 
  
  # Plot the data using ggplot by weekday and weekend
  ggplot(aes(interval, AverageSteps)) + 
  geom_line() + 
  facet_grid(weekday ~ .) +
  
  # and clean the plot up a bit.
  ggtitle('Weekday and Weekend Comparison') +
  xlab('Interval') + 
  ylab('Number of Average Steps')
```